Causal Impact of Masks, Policies, Behavior on Early Covid-19 Pandemic in the U.S.

### 2020-09-09 tex/covidEffectOfNPI-revision.tex
tiny  update

Finally, our causal model is framed using the language of structural equations models and causal diagrams of econometrics (\cite{pwright, haavelmo:1944, tinbergen:1940, wold:1954, pearl:biometrika}) and genetics \citep{wright1923},\footnote{The father and son, P. Wright (economist) and S. Wright (geneticist) collaborated to develop structural equation models and causal path diagrams; P. Wright's key work represented supply-demand system as a directed acyclical graph and established its identification using exclusion restrictions on instrumental variables. We view our work as following this classical tradition.} with natural unfolding potential/structural outcomes representation \citep{rubin1974,tinbergen1930,neyman:PO,imbens_rubin_2015}. The work on causal graphs has been modernized and put on rigorous footing by  \cite{pearl:biometrika} and others (\cite{pearl:robins,pearl:causality, pearl:why,white:chalak,heckman:pinto,hunermund,richardson:mediation,peters2020book,bareinboim:H,hernanrobins2020book}), with major applications taking place especially in computer science and epidemiology. The particular causal diagram we use has several ``mediation" components, where variables affect outcomes directly and indirectly through other variables called mediators; these structures go back at least to \citet[][see Figure 6]{wright1923}; see, e.g., \cite{baron1986}, \cite{Hines2020}, \cite{richardson:mediation} for modern treatments.

@@

The specification of the model above grasps one-period responses.  The dynamics over multiple periods will be induced by the evolution
of information variables, which include time, lagged and integrated values of outcome:\footnote{Our empirical analysis also considers a specification in which information variables include lagged national cases/deaths as well as lagged behavior variables.}
\begin{equation}\label{eq:I}\tag{I}
 {\icolor I_{it} } := I_t( {\ycolor Y_{it}},  { \icolor  I_{i, t-\ell} }) := \Big (g(t) , {\ycolor Y_{it}},  \sum_{m=0}^{ \lfloor t/\ell \rfloor}
{\ycolor Y_{i,t - \ell m}} \Big )'  \end{equation}
 for  each  $t \in \{0,1,...,T\}$, where $g$ is deterministic function of time, e.g., month indicators,
 assuming that the log of new cases at time $ - t \ \mathrm{mod} \ \ell $ is zero, for notational convenience.
 \footnote{The general formula  for  $I_{i, t-1}$ is $
S_{i, t, \ell } + \sum_{m=1}^{ \lfloor t/\ell \rfloor} Y_{i,t - \ell m}$, where $ S_{i, t, \ell }$ is the initial condition, the log of new cases at time $- t \ \mathrm{ mod } \ \ell$.} In this structure, people respond to both global information, captured by a function of time such as month dummies, and local information sources, captured by the local growth rate and the total number of cases. The local information also captures the persistence of the growth rate process.  We model the reaction of people's behavior via the term ${\icolor \gamma'I_{t}}$ in the behavior equation.  The  lagged values of behavior variable may be also included in the information set, but we postpone this discussion after the main empirical results are presented.

>>>>

Finally, our causal model is framed using the language of structural equations models and causal diagrams of econometrics (\cite{pwright, haavelmo:1944, tinbergen:1940, wold:1954, pearl:biometrika}) and genetics \citep{wright1923},\footnote{The father and son, P. Wright (economist) and S. Wright (geneticist) collaborated to develop structural equation models and causal path diagrams; P. Wright's key work represented supply-demand system as a directed acyclical graph and established its identification using exclusion restrictions on instrumental variables. We view our work as following this classical tradition.} with natural unfolding potential/structural outcomes representation \citep{rubin1974,tinbergen1930,neyman:PO,imbens_rubin_2015}. The work on causal graphs has been modernized and developed by \cite{pearl:biometrika,pearl:robins,pearl:causality,pearl:why} and many others (e.g., \cite{pearl:why,white:chalak,richardson:mediation,peters2020book,bareinboim:H,hernanrobins2020book}), with applications in computer science, genetics, epidemiology, and econometrics (see, e.g., \cite{heckman:pinto,hunermund,white:chalak} for applications in econometrics). The particular causal diagram we use has several ``mediation" components, where variables affect outcomes directly and indirectly through other variables called mediators; these structures go back at least to \citet[][see Figure 6]{wright1923}; see, e.g., \cite{baron1986}, \cite{Hines2020}, \cite{richardson:mediation} for modern treatments.

@@

The specification of the model above grasps one-period responses.  The dynamics over multiple periods will be induced by the evolution
of information variables, which include time, lagged and integrated values of outcome:\footnote{Our empirical analysis also considers a specification in which information variables include lagged national cases/deaths as well as lagged behavior variables.}
\begin{equation}\label{eq:I}\tag{I}
 {\icolor I_{it} } := I_t( {\ycolor Y_{it}},  { \icolor  I_{i, t-\ell} }) := \Big (g(t) , {\ycolor Y_{it}},  \sum_{m=0}^{ \lfloor t/\ell \rfloor}
{\ycolor Y_{i,t - \ell m}} \Big )'  \end{equation}
 for  each  $t \in \{0,1,...,T\}$, where $g$ is deterministic function of time, e.g., month indicators,
 assuming that the log of new cases at time $t \leq 0$ is zero, for notational convenience.
 \footnote{The general formula  for  $I_{i, t-1}$ is $
S_{i, t, \ell } + \sum_{m=1}^{ \lfloor t/\ell \rfloor} Y_{i,t - \ell m}$, where $ S_{i, t, \ell }$ is the initial condition, the log of new cases at time $- t \ \mathrm{ mod } \ \ell$.} In this structure, people respond to both global information, captured by a function of time such as month dummies, and local information sources, captured by the local growth rate and the total number of cases. The local information also captures the persistence of the growth rate process.  We model the reaction of people's behavior via the term ${\icolor \gamma'I_{t}}$ in the behavior equation.  The  lagged values of behavior variable may be also included in the information set, but we postpone this discussion after the main empirical results are presented.
